---
title: "Applied Statistical Programming - Vectors and Functions: Messi Lee, Alma Velazquez, Jordan Duffin Wong"
date: "1/24/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\textbf{Write the R code to answer the following questions. Write the code and then
show what the computer returns when that code is run. Make sure to change the appropriate header in the R code block to make this document compile.}

Please put your name at the top of this sheet of paper. You have until the beginning of class 1/26 at 10:00am to answer all of the questions below. You may use R, but not any online documentation. Submit the Rmarkdown and the knitted PDF to Canvas.

1. $e^2$
```{r}
#Code
exp(2)
```

2. $((4)^5)^{\frac{1}{8}}$
```{r}
#Code
((4)^5)^(1/8)
```

3. $sin(\frac{\pi}{3}) \times (1+tan(\frac{\pi}{3}))$
```{r}
#Code
sin(pi / 3) * (1 + tan(pi / 3))
```

4. $\sqrt{14^3 - 6^{\frac{3}{2}}}$
```{r}
#Code
sqrt(14^3 - (6^(3/2)))
```

5. $\vert -ln(2\pi \times (\sqrt{e^9}))\vert$
```{r}
#Code
abs(- (log(2 * pi * sqrt(exp(9)))))
```

6. $\sum\limits^{50}_{i=5}i^{i-1}$
```{r}
#Code
x <- c(5:50)
sum(x^(x-1))
```

7. $\forall i \in [1,50], \text{ } \sqrt{(i)}$
```{r}
#Code
y <- c(1:50)
sqrt(y)
```